\mbox{Specifying Imperative Higher Order Programs}

### 2009-10-01 papers/esop2010/main.tex
ok

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\lift}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\correct}[1]{ c(#1) }

@@

We now proceed to explain how proof obligations are obtained from the
annotated program.

In the following, we use the metasymbol $v$ to denote well-typed values, i.e.
terms with a typing derivation $Γ|-_v v : τ$.

@@

\begin{figure}[htbp]
  \begin{eqnarray*}
    c(x~[\alist{κ}]) &=& \Trueml \\
    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
    c(t~t') &=& c(t) /\ c(t') \\
    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
    ∀s:\efft{ε}.p~s => \wpre_s(t,q \\
    \wpre_s(v, q) &=& q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : τ, ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.λx:\ceil{τ}.\wpre_{s'}(t2, q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

>>>>

\newcommand{\wpre}{{\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{\langle #1 \rangle}
\newcommand{\lift}[1]{\langle #1 \rangle}
\newcommand{\alist}[1]{\overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\correct}[1]{c(#1)}

@@

The aim of the weakest precondition calculus is to obtain proof obligations
from the annotated program. To do this, the calculus starts from a formula
which holds, or is supposed to hold, at the end of the program, and walks
backwards in the program source to obtain the weakest formula which has to
hold at the beginning of the program (hence the name {\em weakest
precondition}.

In the following, we use the metasymbol $v$ to denote well-typed values, i.e.
terms with a typing derivation $Γ|-_v v : τ$. In the definition of the weakest
precondition calculus, we will use the notation $(t : τ, ε)$ to denote that
$t$ has type $τ$ and effect $ε$. We assume that applications of the
subeffecting rule {\sc Sub} rule are explicitely inserted in the term; they
are written $(t:ε<:ε')$.

@@

First of all, analogous to the lifting of types to the logic level, we define
a lifting of values to the logic level. Similarly, this translation traverses
the value until it encounters an impure function. From the logic point of
view, an impure function is represented by its specification, so the
translation keeps only the pair of pre- and postcondition, both abstracted
over the function argument. The function body is thrown away.

\begin{figure}[htbp]
  \begin{eqnarray*}
    c(x~[\alist{κ}]) &=& \Trueml \\
    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
    c(t~t') &=& c(t) /\ c(t') \\
    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
    ∀s:\efft{ε}.p~s => \wpre_s(t,q~s) \\
    \wpre_s(v, q) &=& c(v) /\ q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct{t'} /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      ∀\alist{χ}.c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : τ,ε) ~\inml~ t2 , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε }.λx:\ceil{τ}.\wpre_{s'}( t2 , q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

The weakest precondition calculus itself comes in two parts, one for values
and one for effectful terms. Both are reasonably straightforward. The {\em
correctness} of values, written $c(v)$, ensures that any effectful function
appearing in a value adheres to its specification. To this end, the value is
traversed, quantifying over bound variables and using conjunction for pure
applications. When an effectful function is encountered, one has to prove that
the precondition of the function implies the postcondition, given the function
body.

The part concerning effectful functions, denoted $\wpre_s(t,q)$ takes as
argument a term $t$ of type $τ$ and effect $ε$, a postcondition $q$ of type
$\efft{ε} -> τ -> \propml$ and a current state $s$ of type $\efft{ε}$. If the
term is in fact a value, the value has to be correct and the postcondition
must be true for the lifted value and the current state. In the case of an
application, both values must be correct, the precondition of the function
must hold in the current state, and for any state and return value of the
function, the function's postcondition must imply the condition $q$ to be
proved.

As for the typing relation, we have two rules for the \letml-binding. If the
bound term is a value, we can translate the \letml-binding in the program by a
\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
the argument of the \wpre of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
t2 $). The rule for \ifml-expressions is straightforward. The rule for
subeffecting adapts the domain of the current store and the expected domain of
the postcondition so that the types work out.
